---
title: "Uber Pickup Data Analysis"
author: "Debalina Chakraborty & Samantha Lobo"
date: "4/18/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# **Motivation & Overview**
With the increasing amount of car share services, we have seen that there have been a lot of customer dissatisfaction leading to poor company reviews and customer experiences.
Our objective is to draw insights from the historical Uber Pickup data available and suggest solutions which will further enhance customer experience.
This is a predominantly data visualization project using the ggplot2 library for understanding the data and for developing a story to further understand the customers who avail the trips in New York. 

# **Related work**
Previous work has been done to predict customer behaviour based on Uber pickups & determining the rush hours. We have used the base idea of this project and have performed exploratory data analysis based on visualizations only.
a(Https://www.analyticsvidhya.com/blog/2021/10/end-to-end-predictive-analysis-on-ubers-data/)

# ** Initial questions**
The questions we try to answer in this project are:
1. How do passengers fare throughout the day?
2. Which part of the day do we have more number of trips?
3. Which day, month contributed to the highest trips in the year?
4. How passengers made trips from different bases?

>Through the course of this project, we evolve into questions such as:
How has time affected the customer trips?

## R Markdown
```{r}
###### Loading the libraries
library(ggplot2)
library(ggmap)
library(plotly)
library(ggthemes)
library(lubridate)
library(dplyr)
library(tidyr)
library(DT)
library(scales)
library(leaflet)
```

#**Data**
###### Collecting the data : The data was collected from Kaggle competition a('https://www.kaggle.com/datasets/fivethirtyeight/uber-pickups-in-new-york-city'). We are loading the data sets which are individual files of pickup data from april 2014 to September 2014 and combine them to create a data set for the whole year of 2014.

```{r}

###Create a vector of our colors that will be included in our plotting functions
colors = c("#E6E6FA", "#D8BFD8", "#DDA0DD", "#EE82EE", "#DA70D6", "#BA55D3", "#8A2BE2")
getwd()

###### Load the data & data wrangling, storing them in dataframes to be used later.
apr_data <- read.csv("uber-raw-data-apr14.csv")
may_data <- read.csv("uber-raw-data-may14.csv")
jun_data <- read.csv("uber-raw-data-jun14.csv")
jul_data <- read.csv("uber-raw-data-jul14.csv")
aug_data <- read.csv("uber-raw-data-aug14.csv")
sep_data <- read.csv("uber-raw-data-sep14.csv")
data_2015 <- read.csv("uber-raw-data-janjune-15.csv")
data_2014 <- rbind(apr_data,may_data, jun_data, jul_data, aug_data, sep_data)

data_2014$Date.Time <- as.POSIXct(data_2014$Date.Time, format = "%m/%d/%Y %H:%M:%S")

data_2014$Time <- format(as.POSIXct(data_2014$Date.Time, format = "%m/%d/%Y %H:%M:%S"), format="%H:%M:%S")

data_2014$Date.Time <- ymd_hms(data_2014$Date.Time)

data_2014$day <- factor(day(data_2014$Date.Time))
data_2014$month <- factor(month(data_2014$Date.Time, label = TRUE))
data_2014$year <- factor(year(data_2014$Date.Time))
data_2014$dayofweek <- factor(wday(data_2014$Date.Time, label = TRUE))

data_2014$hour <- factor(hour(hms(data_2014$Time)))
data_2014$minute <- factor(minute(hms(data_2014$Time)))
data_2014$second <- factor(second(hms(data_2014$Time)))
data_2014$Base_names <- if_else(data_2014$Base == "B02512", "Unter", 
                                if_else(data_2014$Base == "B02598", "Hinter", 
                                if_else(data_2014$Base == "B02682", "Schmecken",
                                if_else(data_2014$Base == "B02764", "Danach-NY", 
                                if_else(data_2014$Base == "B02617", "Weiter",
                                if_else(data_2014$Base == "B02765", "Grun",
                                if_else(data_2014$Base == "B02835", "Dreist",
                                if_else(data_2014$Base == "B02836", "Drinnen",data_2014$Base))))))))
data_2015$Base_names <- if_else(data_2015$Affiliated_base_num == "B02512", "Unter", 
                                if_else(data_2015$Affiliated_base_num == "B02598", "Hinter", 
                                if_else(data_2015$Affiliated_base_num == "B02598", "Hinter", 
                                if_else(data_2015$Affiliated_base_num == "B02682", "Schmecken",
                                if_else(data_2015$Affiliated_base_num == "B02764", "Danach-NY", 
                                if_else(data_2014$Base == "B02617", "Weiter",
                                if_else(data_2015$Affiliated_base_num == "B02765", "Grun",
                                if_else(data_2015$Affiliated_base_num == "B02835", "Dreist",
                                if_else(data_2015$Affiliated_base_num == "B02836",  "Drinnen",data_2015$Affiliated_base_num)))))))))
head(data_2014)
```

#**Exploratory Data Analysis**

###### Below graph plotting highlights the trips by the hours in a day in the year 2014
```{r}
 hour_data <- data_2014 %>%
  group_by(hour) %>%
  dplyr::summarize(Total = n()) 
datatable(hour_data)

Trip_p_hr<- ggplot(hour_data, aes(hour, Total)) + 
  geom_bar( stat = "identity", fill = "purple", color = "black") +
  ggtitle("Trips Every Hour") +
  theme(legend.position = "none") +
  scale_y_continuous(labels = comma)
ggplotly(Trip_p_hr)

```

###### Below graph plotting highlights the trips by the hours for every month in the year 2014
```{r}

month_hour <- data_2014 %>%
  group_by(month, hour) %>%
  dplyr::summarize(Total = n())

Trip_p_hr_month <- ggplot(month_hour, aes(hour, Total, fill = month)) + 
  geom_bar( stat = "identity") +
  ggtitle("Trips by Hour and Month") +
  scale_y_continuous(labels = comma)+
  scale_fill_manual(values = colors)
ggplotly(Trip_p_hr_month)
```

###### This shows the trip every day in a month. We see a general trend that over 15,000 trips were covered on an average between 11-16th of a month & again a small spike towards the end of the month.
```{r}
day_group <- data_2014 %>%
  group_by(day) %>%
  dplyr::summarize(Total = n()) 
datatable(day_group)

Trip_p_day <- ggplot(day_group, aes(day, Total)) + 
  geom_bar( stat = "identity", fill = "purple") +
  ggtitle("Trips Every Day") +
  theme(legend.position = "none") +
  scale_y_continuous(labels = comma)+
  scale_fill_manual(values = colors)
ggplotly(Trip_p_day)
```
###### Let's see how the above daily trips spread across the months, to get a better idea of when we see a peak in Uber pickups.
```{r}
day_month_group <- data_2014 %>%
  group_by(month, day) %>%
  dplyr::summarize(Total = n())

Trip_p_day_month <- ggplot(day_month_group, aes(day, Total, fill = month)) + 
  geom_bar( stat = "identity") +
  ggtitle("Trips by Day and Month") +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = colors)
ggplotly(Trip_p_day_month)

```
###### Here we can see that the highest uber pickups were between July-Sept. Considering that being the tourist season in NYC the numbers are justified.
```{r}
month_group <- data_2014 %>%
  group_by(month) %>%
  dplyr::summarize(Total = n()) 
datatable(month_group)

Trips_p_month <- ggplot( month_group, aes(month, Total, fill = month)) + 
  geom_bar( stat = "identity") +
  ggtitle("Trips by Month") +
  theme(legend.position = "none") +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = colors)
ggplotly(Trips_p_month)

```

###### As we can see, from the below graph, the peak pickup between April - June is during the weekdays, but as we move towards July-September we can see that the highest pickups were during the weekends (Friday-Sat).
```{r}
month_weekday <- data_2014 %>%
  group_by(month, dayofweek) %>%
  dplyr::summarize(Total = n())

Trips_p_day_month <- ggplot(month_weekday, aes(month, Total, fill = dayofweek)) + 
  geom_bar( stat = "identity", position = "dodge") +
  ggtitle("Trips by Day and Month") +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = colors)
ggplotly(Trips_p_day_month)
```


###### We are now analysing Uber pickup by Base locations, most of the pickups happen from NYC bases like Hinter, Weiter & Schmecken
```{r}
Trip_By_Base <- ggplot(data_2014, aes(Base_names)) + 
  geom_bar(fill = "purple") +
  scale_y_continuous(labels = comma) +
  ggtitle("Trips by Bases")
ggplotly(Trip_By_Base)
```

###### Below graph highlights the picups made across these Uber bases in every month. All across the bases the number of trips increased between July & September. Except in Hinter & Schmecken where the count of trips were higher in Aril & May.
```{r}
Trip_By_Base_Month <- ggplot(data_2014, aes(Base_names, fill = month)) + 
  geom_bar(position = "dodge") +
  scale_y_continuous(labels = comma) +
  ggtitle("Trips by Bases and Month") +
  scale_fill_manual(values = colors)
ggplotly(Trip_By_Base_Month)
```

###### We are now analysing Uber pickup by Base locations by week, most of the trips are during weekdays.
```{r}
Trip_ByWeek_Base <- ggplot(data_2014, aes(Base_names, fill = dayofweek)) + 
  geom_bar(position = "dodge") +
  scale_y_continuous(labels = comma) +
  ggtitle("Trips by Bases and DayofWeek") +
  scale_fill_manual(values = colors)
ggplotly(Trip_ByWeek_Base)
```

###### Creating a Heatmap visualization of day, hour and month. 
###### a) The first graph helps us understand that almost everyday  6-8 am or the first 6-8th hour is the peak pickup time for uber rides & similarly the next rush hour is between 4-8 PM that is between 16th & 20th hour of the day.

```{r}

day_and_hour <- data_2014 %>%
  group_by(day, hour) %>%
  dplyr::summarize(Total = n())

datatable(day_and_hour)

ggplot(day_and_hour, aes(day, hour, fill = Total)) +
  geom_tile() +
  scale_fill_gradient(low = "#E6E6FA", high= "#4B0082", guide = "colourbar")+
  ggtitle("Heat Map by Hour and Day")

```
###### b) The second graphs shows the peak uber rides happen in the Month of Aug-Sep.
```{r}

ggplot(day_month_group, aes(day, month, fill = Total)) +
  geom_tile() +
  scale_fill_gradient(low = "#E6E6FA", high= "#4B0082", guide = "colourbar")+
  ggtitle("Heat Map by Month and Day")

```
###### c) The third graphs shows the pickup ride distribution by month & day of the week, where we see a clear distinguishment between the trend till July where max pickups are in Weekdays & then during Septemeber it shifts towards the weekend.
```{r}

ggplot(month_weekday, aes(dayofweek, month, fill = Total)) +
  geom_tile() +
  scale_fill_gradient(low = "#E6E6FA", high= "#4B0082", guide = "colourbar")+
  ggtitle("Heat Map by Month and Day of Week")
```

###### Further diving deeper we see the trend between the pickups in a month by Base locations
```{r}
month_base <-  data_2014 %>%
  group_by(Base_names, month) %>%
  dplyr::summarize(Total = n()) 

day0fweek_bases <-  data_2014 %>%
  group_by(Base_names, dayofweek) %>%
  dplyr::summarize(Total = n()) 

ggplot(month_base, aes(Base_names, month, fill = Total)) +
 geom_tile() +
  scale_fill_gradient(low = "#E6E6FA", high= "#4B0082", guide = "colourbar")+
   ggtitle("Heat Map by Month and Bases")
```

###### And Pickup from a base in every day of a week.
```{r}
ggplot(day0fweek_bases, aes(Base_names, dayofweek, fill = Total)) +
  geom_tile() +
  scale_fill_gradient(low = "#E6E6FA", high= "#4B0082", guide = "colourbar")+
  ggtitle("Heat Map by Bases and Day of Week")
```


###### Analysing the Top Pickup locations in NYC - We can see from the graphh that the top pickup location are JFK International airport, LaGuardia Airport, TWA Hotels near JFK, LaGuardia delta Airlines Terminal, 9/11 Museum & LaGuardia Southwest airline terminal.
```{r}
uber_2014_top <- data_2014 %>%
  mutate(Lat_3 = round(Lat, 3),Lon_3 = round(Lon, 3)) %>% 
  count(Lat_3, Lon_3, sort = TRUE) %>% 
  head()

uber_2014_top_map <- leaflet(uber_2014_top)
uber_2014_top_map <- addTiles(uber_2014_top_map)
uber_2014_top_map <- addCircleMarkers(uber_2014_top_map, lng = ~Lon_3,
                                        lat = ~Lat_3)
uber_2014_top_map
```


#**Narrative & summary**

1. Maximum number of trips occur between 16th to 19th hour of the day which is approximately between 4PM to 7 Pm.
2. The total pickups are fewer in the months of April to Jun but increases significantly from July to September, which is the preferred months for Toursist to vist NYC.
3. We see a general trend that over 15,000 trips were covered on an average between 11-16th of a month & again a small spike towards the end of the month.
4. When we compare the data between the days in a week we see that between April - June most pickups are during the weekdays, but as we move towards July-September we can see that the highest pickups were during the weekends (Friday-Sat).
5. Further analysing Uber pickup by Base locations, most of the pickups happen from NYC bases like Hinter, Weiter & Schmecken.
6. Below are the top pickup locations NYC which contribute to maximum rides. 
	- JFK International airport
	- LaGuardia Airport
	- TWA Hotels near JFK
	- LaGuardia delta Airlines Terminal
	- 9/11 Museum 
	- LaGuardia Southwest airline terminal.
	
We can conclude that 50% of the rides are available during 4-8 PM within the city & July-September being the preferred months of travel for tourists, if we can make Uber rides available near the Top 5 locations at max capacity Uber might be able to reduce their customer dissatisfaction scores.
